Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-02-22 src/tex/ms.tex
improved reweighted posteriors

Eq.~\eqref{eq:waveform_modification} can be derived as the first order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from a truncated series expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
\left|\kappa\right| \left(d_C/1\,\mathrm{Gpc}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, .
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

@@

\subsubsection{Nonshared birefringence parameters}

@@

Fig.~\ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}.
\begin{table}
    \caption{Events that best constrain $\kappa$, where $\tilde{\kappa}$ is the posterior median of $\kappa$.} 
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

@@

Another set of events stands out in Fig.~\ref{fig:violin_kappa} due to evident bimodality in the $\kappa$ posterior.
That is the case for the events listed in Table~\ref{tab:bimodal_events_mass}.
\begin{table}
    \caption{Events that show bimodality in the $\kappa$ posterior, where the total mass of the binary is given in solar masses.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}
We understand this effect to be a consequence of the narrow sets of frequencies that are visible for these heavy events, as we mentioned in Sec.~\ref{sec:inclination} and further demonstrate below.

@@

Fig.~\ref{fig:violin_kappa} shows some variance in the distribution of $\kappa$ posteriors, including some apparent outliers like GW170818.
This is not unexpected: assuming independent Gaussian noise, we might expect up to ${\sim}3$ events in a sample of 71 events to deviate from the true value of $\kappa$ by ${\sim}2\sigma$.
\mi{WIP}

% Corner plot of the mean and standard deviation of kappa
We then show the result of Bayesian hierarchical modeling in Fig.~\ref{fig:corner_Gaussian}.
The median of $\mu$ within $90\%$ confidence interval is \variable{output/mu_median.txt}.
The median of $\sigma$ within $90\%$ confidence interval is \variable{output/sigma_median.txt}.
The posterior of $\mu$ is concentrated around $0$, which means the population constraint on $\kappa$ is consistent with GR.

@@

@@

\section{Reweighed kappa distributions}
\label{sec:reweighed_kappa}

\begin{figure}
    \script{reweighed_kappa.py}
    \includegraphics[width=\columnwidth]{figures/reweighed_kappa.pdf}
    \caption{Reweighed $\kappa$ distributions.}
    \label{fig:reweighed_kappa}
\end{figure}

>>>>

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from a truncated series expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
\left|\kappa\right| \left(d_C/1\,\mathrm{Gpc}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, .
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

@@

\subsubsection{Nonshared birefringence parameters}
\label{sec:method:hier}

@@

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}.
\begin{table}
    \caption{Events that best constrain $\kappa$, where $\tilde{\kappa}$ is the posterior median of $\kappa$.} 
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

@@

Another set of events stands out in Fig.~\ref{fig:violin_kappa} due to evident bimodality in the $\kappa$ posterior.
That is the case for the events listed in Table~\ref{tab:bimodal_events_mass}.

\begin{table}
    \caption{Events that show bimodality in the $\kappa$ posterior, where the total mass of the binary is given in solar masses.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}
We understand this effect to be a consequence of the narrow sets of frequencies that are visible for these heavy events, as we mentioned in Sec.~\ref{sec:inclination} and further demonstrate below.

@@

Figure \ref{fig:violin_kappa} shows a degree of variance in the distribution of $\kappa$ posteriors for different events, including some apparent outliers like GW170818.
This is not unexpected: assuming independent Gaussian noise for each event, we might expect up to ${\sim}3$ out of the 71 posteriors (i.e., ${\sim}5\%$) to deviate away from the true $\kappa$ value by over ${\sim}2\sigma$.

To further assess the statistical properties of the set of posteriors in Fig.~\ref{fig:violin_kappa}, we apply the hierarchical analysis described in Sec.~\ref{sec:method:hier}.
By characterizing the population mean and standard deviation over $\kappa$, this also allows us to obtain a collective measurement that does not assume all events share the same value of $\kappa$.

@@

We summarize the result of this exercise in Fig.~\ref{fig:corner_Gaussian}, which shows the posterior on the population mean $\mu$ and standard deviation $\sigma$ inferred from the collection of observations in Fig.~\ref{fig:violin_kappa}.
This posterior supports lack of birefringence ($\mu = \sigma = 0$) within 90\% credibility; from the marginals, we constrain $\mu =$ \variable{output/mu_median.txt} and $\sigma =$ \variable{output/sigma_median.txt}, for 90\%-credible symmetric intervals around the median \mi{THOMAS: for $\sigma$ it would make more sense to report the one-sided 90\%-credible upper limit here, i.e.i, the 90th percentile, as $\sigma < XX$}.
However, even though $\sigma = 0$ is well supported, the $\sigma$ posterior peaks away from the origin, indicating some preference for a nonzero variance.
This could be a sign of the presence of outliers in our sample.

As a check for outliers, we reconsider the set of measurements in Fig.~\ref{fig:violin_kappa} in light of the hierarchical result for $\mu$ and $\sigma$ in Fig.~\ref{fig:corner_Gaussian}.
This amounts to reweighting the $\kappa$ posterior for each event under a population prior marginalized over $\mu$ and $\sigma$, conditional on the measurements from all other events \cite{Callister:T2100301}.

\begin{figure}
    \script{reweighed_kappa.py}
    \includegraphics[width=\columnwidth]{figures/reweighed_kappa.pdf}
    \caption{Reweighted $\kappa$ distributions.}
    \label{fig:reweighted_kappa}
\end{figure}

@@
